Sufficient Conditions for Reachability in Automata Networks with Priorities

### 2014-04-07 main.tex

% Commandes À FAIRE
\usepackage{color} % Couleurs du texte
%\definecolor{darkgreen}{rgb}{0,0.5,0}
%\newcommand{\towrite}[1]{\textcolor{darkgreen}{[#1]}}
\newcommand{\todo}[1]{\textcolor{red}{\textbf{[#1]}}}

@@

To address this issue, we propose an extended definition of the framework, including priority classes for actions.
In this paper, we focus on a restriction of the Process Hitting with two classes of priorities and a specific behaviour of the components, that is sufficient to tackle the aforementioned problem of cooperations.
We show that this class of Process Hitting models allows to represent any
Asynchronous Discrete Networks, either Boolean or multivalued; and also any
Process Hitting model with arbirtrary number of priorty classes.
Then we develop a new refinement for the under-approximation of the static analysis to give accurate results for this class of Process Hitting models.
Our method thus allows to efficiently under-approximate reachability properties
in Asynchronous Discrete Networks,
as it is illustrated on the model-checking of a signaling network of 94
components, which is unprecedented.
\end{abstract}
\begin{keyword}
discrete networks \sep
abstract interpretation \sep
reachability \sep
qualitative models \sep
systems biology
\end{keyword}
\end{frontmatter}

>>>>

% Commandes À FAIRE
\usepackage{color} % Couleurs du texte
%\definecolor{darkgreen}{rgb}{0,0.5,0}
%\newcommand{\towrite}[1]{\textcolor{darkgreen}{[#1]}}
%\newcommand{\todo}[1]{\textcolor{red}{\textbf{[#1]}}}

@@

To address this issue, we propose an extended definition of the framework, including priority classes for actions.
In this paper, we focus on a restriction of the Process Hitting with two classes of priorities and a specific behaviour of the components, that is sufficient to tackle the aforementioned problem of cooperations.
We show that this class of Process Hitting models allows to represent any
Asynchronous Discrete Networks, either Boolean or multivalued; and also any
Process Hitting model with arbitrary number of priority classes.
Then we develop a new refinement for the under-approximation of the static analysis to give accurate results for this class of Process Hitting models.
Our method thus allows to efficiently under-approximate reachability properties
in Asynchronous Discrete Networks,
as it is illustrated on the model-checking of a signalling network of 94
components, which is unprecedented.
\end{abstract}
\begin{keyword}
discrete networks \sep
abstract interpretation \sep
reachability \sep
qualitative models \sep
systems biology
\end{keyword}
\end{frontmatter}

### 2014-04-07 parts/def-ph.tex

  \noindent
  Furthermore, if $\mysigma \in \PHsubl[\PHl]$ and $s \in \PHl$, we note:
    \[\mysigma \subseteq s \EQDEF \forall a_i \in \Proc, a_i \in \mysigma \Rightarrow a_i \in s \enspace.\]
\end{definition}
%
\begin{definition}[$\Cap : \PHl \times \PHproc \rightarrow \PHl$]
\label{def:statecap}
  Given a state $s \in \PHl$ and a process $a_i \in \PHproc$, $(s \Cap a_i)$ is the state defined by:
  $\PHget{(s \Cap a_i)}{a} = a_i \wedge \forall b \neq a, \PHget{(s \Cap a_i)}{b} = \PHget{s}{b}$.
  We also extend this definition to a set of processes given that all processes are from different sorts by the override of each process:
  $\forall ps \in \PHsubl[\PHl], s \Cap \toset{ps} = s \underset{a_i \in \toset{ps}}{\Cap} a_i$.
  %$\forall as \subseteq \PHs, \forall ps \in \underset{a \in as}{\times} \PHl_a, s \Cap ps = s \underset{a_i \in ps}{\Cap} a_i$.
\end{definition}
%
\begin{definition}[Dynamics of a PH ($\PHPtrans$)]
\label{def:play}
  An action $h = \PHhit{a_i}{b_j}{b_k} \in \PHa^{(n)}$ of priority $n$ is \emph{playable} in $s \in \PHl$
  if and only if $\PHget{s}{a} = a_i$, $\PHget{s}{b} = b_j$ and $\forall m < n, \forall g \in \PHa^{(m)}, \PHhitter(g) \notin s \vee \PHtarget(g) \notin s$.
  In such a case, $(s \PHplay h)$ stands for the state resulting from the play of the action $h$ in $s$ and is defined by: $(s \PHplay h) = s \Cap b_k$.
  Moreover, we denote: $s \PHPtrans (s \PHplay h)$.

@@

In this subsection, we give several general lemmas that can be derived from the restrictions of \pref{ssec:hypothesis}, %(\pref{cr:bounded}, \ref{cr:cyclefreeness} \& \ref{cr:compcs})
in the special case of a PH model with $2$ classes of priorities $\PH = (\PHs; \PHl; \PHa^{\langle 2 \rangle})$.
These results will help building the static analysis of \pref{sec:sa}.

>>>>

  \noindent
  Furthermore, if $\mysigma \in \PHsubl[\PHl]$ and $s \in \PHl$, we note:
    \[\mysigma \subseteq s \EQDEF \forall a_i \in \Proc, a_i \in \mysigma \Rightarrow a_i \in s \enspace.\]
\end{definition}
%
\begin{definition}[$\Cap : \PHl \times \PHproc \rightarrow \PHl$]
\label{def:statecap}
  Given a state $s \in \PHl$ and a process $a_i \in \PHproc$, $(s \Cap a_i)$ is the state defined by:
  $\PHget{(s \Cap a_i)}{a} = a_i \wedge \forall b \neq a, \PHget{(s \Cap a_i)}{b} = \PHget{s}{b}$.
  We also extend this definition to a set of processes given that all processes are from different sorts by the override of each process:
  $\forall ps \in \PHsubl[\PHl], s \Cap \toset{ps} = s \underset{a_i \in \toset{ps}}{\Cap} a_i$.
\end{definition}
%
\begin{definition}[Dynamics of a PH ($\PHPtrans$)]
\label{def:play}
  An action $h = \PHhit{a_i}{b_j}{b_k} \in \PHa^{(n)}$ of priority $n$ is \emph{playable} in $s \in \PHl$
  if and only if $\PHget{s}{a} = a_i$, $\PHget{s}{b} = b_j$ and $\forall m < n, \forall g \in \PHa^{(m)}, \PHhitter(g) \notin s \vee \PHtarget(g) \notin s$.
  In such a case, $(s \PHplay h)$ stands for the state resulting from the play of the action $h$ in $s$ and is defined by: $(s \PHplay h) = s \Cap b_k$.
  Moreover, we denote: $s \PHPtrans (s \PHplay h)$.

@@

In this subsection, we give several general lemmas that can be derived from the restrictions of \pref{ssec:hypothesis},
in the special case of a PH model with $2$ classes of priorities $\PH = (\PHs; \PHl; \PHa^{\langle 2 \rangle})$.
These results will help building the static analysis of \pref{sec:sa}.

### 2014-04-07 parts/suppl-ph-dn.tex

% vim:set spell spelllang=en:

\section{Weak Bisimulation of Asynchronous Discrete Networks (\pref{ssec:hypothesis})}
\label{suppl:demotrad}

We exhibit an encoding of Asynchronous Discrete Networks (ADN) with the Process
Hitting using two classes of priorities, and prove a weak bisimulation relation.

A Discrete Network gathers a finite number of components $i\in\segm{1}{n}$ having a discrete finite domain
$\mathbb F^i$ that we note $\mathbb{F}^i = \segm{1}{l_i}$.
For each component $i\in\segm{1}{n}$, a map $\mathbb F \rightarrow \mathbb F^i$ is defined, where
$\mathbb F = \mathbb F^1 \times \cdots \times \mathbb F^n$, giving the next value of the component
with respect to the global state of the network.
Typically $f^i$ depends on a subset of components that we denote $\DNdep(f^i)$.
In the case of Asynchronous Discrete Networks (ADN), a transition relation $\DNtrans\subseteq \mathbb
F\times \mathbb F$ is defined such that $x\DNtrans x'$ if and only if there exists a unique
$i\in\segm{1}{n}$ such that $\get{x'}{i}=f^i(x)$ and $\forall j\in\segm{1}{n}, j\neq i, \get{x'}{j}
=\get{x}{j}$, \ie one and only one component has been updated.
This is formalised in \pref{def:DN}.

\begin{definition}[Asynchronous Discrete Network (ADN)]
\label{def:DN}
  An ADN is defined by a couple $(\mathbb F, \langle f^1, \dots, f^n \rangle)$
  where $\mathbb{F} = \mathbb{F}^1\times\dots\times\mathbb{F}^n$,
  and $\forall i\in\segm{1}{n}$,
  $f^i: \mathbb{F} \rightarrow \mathbb{F}^i$ with
  $\mathbb{F}^i = \segm{1}{l_i}$.
  Given two states $x,x'\in\mathbb F$, the transition relation $\DNtrans$ is given by
  \[
  x\DNtrans x' \Longleftrightarrow
    \exists i\in\segm{1}{n}, f^i(x)=\get{x'}{i}
    \wedge \forall j\in\segm{1}{n}, j\neq i, \get{x}{j}=\get{x'}{j}
  \enspace,
  \]
  where $\get{x}{i}$ is the $i$-th component of $x$.
  %
  We note $\DNdep(f^i)\subseteq \segm{1}{n}$ the set of components on which the value of $f^i$
  depends: $\forall x,x'\in \mathbb F$ such that $\forall
  j\in\DNdep(f^i), \get{x}{j}=\get{x'}{j}$, necessarily $f^i(x)=f^i(x')$.
\end{definition}

Let us denote the encoding of the ADN $\DNdef$ in Process Hitting with $2$ classes
of priorities by $\toPH\DNdef$ (\pref{def:DN2PH}).
For each component $i\in\segm{1}{n}$ of the ADN, two sorts are built: $a^i$ acting for the component
value, and $f^i$ acting for a cooperative sort between the components $\DNdep(f^i)$.
Sorts $a^i$ have one process $a^i_k$ per element in $k\in\mathbb F^i$.
Sorts $f^i$ have one process $f^i_\varsigma$ per state $\varsigma \in \underset{j\in\DNdep(f^i)}{\times}
\PHl_{a^j}$.
Two classes of actions are then defined:
$\Hits^{(1)}$ is the set of actions updating the cooperative sorts according to the current state of the
components:
if $j\in\DNdep(f^i)$, $a^j_k$ hits each process $f^i_\varsigma$ where $\get{\varsigma}{a^j}\neq
a^j_k$ to make it bounce to the process $f^i_{\varsigma \Cap a^j_k}$.
%$f^i_{\varsigma'}$ where $\get{\varsigma'}{a^j}=a^j_k$.
$\Hits^{(2)}$ is the set of actions encoding the transitions in the ADN:
$f^i_\varsigma$ hits the processes of sort $a^i$ to make them bounce to the process
$a^i_{k'}$ if and only if $k'=f^i(\decode \varsigma)$;
$\decode \varsigma$ being the ADN state corresponding to the PH (partial) state $\varsigma$ (note that
$f^i(\decode \varsigma)$ is fully defined because $\decode \varsigma$ specifies the state for all
the components in $\DNdep(f^i)$).

\begin{definition}
\label{def:DN2PH}
  $\toPH\DNdef=(\Sigma,\PHl,(\Hits^{(1)},\Hits^{(2)}))$ is the Process Hitting with 2 classes of priority encoding the
  ADN $\DNdef$, with:
  \begin{itemize}
    \item $\Sigma = \{ a^1, \dots, a^n \} \cup \{ f^1, \dots, f^n \}$,
      the sorts for components ($a^i$) and cooperative sorts ($f^i$);

    \item $\PHl=\underset{i\in\segm{1}{n}}{\times} \PHl_{a^i} \times \underset{i\in\segm{1}{n}}{\times} \PHl_{f^i}$, where
      $\PHl_{a^i}=\{a^i_0, \dots, a^i_{l_i}\}$, and
      $\PHl_{f^i}=\{f^i_\varsigma \mid \varsigma\in\underset{j\in\DNdep(f^i)}{\times} \PHl_{a^i} \}$
      if $\DNdep(f^i)\neq\emptyset$, otherwise
      $\PHl_{f^i}=\{f^i_\emptyset\}$;

    \item $\Hits^{(1)}= \{ \hit{a^j_k}{f^i_\varsigma}{f^i_{\varsigma'}}
      \mid i\in\segm{1}{n} \wedge
      j\in\DNdep(f^i) \wedge a^j_k\in \PHl_{a^j} \wedge f^i_\varsigma\in \PHl_{f^i}
      \wedge
      \get{\varsigma}{a^j}\neq a^j_k 
      \wedge \get{\varsigma'}{a^j}=a^j_k
      \wedge (\get{\varsigma'}{a^{l}}=\get{\varsigma}{a^l},
      \forall l\in\segm{1}{n},l\neq j)
      \}$, 
      the set of actions with priority $1$ that update cooperative sorts;

    \item $\Hits^{(2)}=\{ \hit{f^i_\varsigma}{a^i_k}{a^i_{k'}} \mid i\in\segm{1}{n} \wedge
          f^i_\varsigma\in \PHl_{f^i} \wedge
          a^i_k\in \PHl_{a^i}\wedge k\neq k' \wedge f^i(\decode \varsigma) = k'
        \}$,
      the set of actions with priority $2$ for updating the components using their respective discrete
      maps.
      $\decode \varsigma$ is defined below.
  \end{itemize}
  Given a state $s\in \PHl$ of the Process Hitting, 
  $\decode s=x$ is the corresponding state in the ADN:
  $\forall i\in\segm{1}{n}, \get{s}{a^i}=a^i_k \Rightarrow \get{x}{i}=k$.

  \noindent
  Given a state $x\in \mathbb F$ of the ADN, 
  $\encode x=s$ is the corresponding state in the Process Hitting:
  $\forall i\in\segm{1}{n}, \get{x}{i}=k \Rightarrow \get{s}{a^i}=a^i_k$
  and
  $\forall i\in\segm{1}{n}, \get{s}{f^i}=f^i_\varsigma$ with $f^i_\varsigma\in \PHl_{f^i}$
  and $\forall j\in\DNdep(f^i), \get{\varsigma}{j}=\get{s}{a^j}$.
\end{definition}

\pref{thm:bisimDN} states the weak bisimulation relation between an ADN and its encoding in
PH with $2$ classes of priorities.
Intuitively, actions updating cooperative sorts being prioritised, actions updating component
sorts follow strictly the possible transitions of the ADN.

\begin{theorem}[$\DNdef \approx \toPH\DNdef$]~
\label{thm:bisimDN}
  \begin{enumerate}
    \item \label{adn2ph} $\forall x,x'\in\mathbb F$,
      $x\DNtrans x' \Longrightarrow \encode x \PHPtrans^* \encode{x'}$,
      where $\PHPtrans^*$ is a finite sequence of $\PHPtrans$ transitions.

    \item \label{ph2adn} $\forall s,s'\in \PHl$,
      $s\PHPtrans s' \Longrightarrow \decode s = \decode {s'} \vee \decode s \DNtrans
      \decode{s'}\enspace.$
  \end{enumerate}
\end{theorem}
\begin{proof}
(\ref{adn2ph}) From \pref{def:DN}, $x\DNtrans x'\Rightarrow \exists i\in\segm{1}{n},
f^i(x)=\get{x'}{i} \wedge \forall j\in\segm{1}{n},i\neq j, \get{x}{j}=\get{x'}{j}$.
Let us assume (without loss of generality) that $f^i(x)=k'$, $\get{x}{i}=k$ and
$\varsigma\in\underset{j\in\DNdep(f^i)}{\times} \PHl_{a^j}$ such that
$\forall j\in\DNdep(f^i), \get{\varsigma}{j}=a^j_{\get{x}{j}}$.
From \pref{def:DN2PH}, $h=\hit{f^i_\varsigma}{a^i_k}{a^i_{k'}}\in\Hits^{(2)}$.
From the definition of $\encode x$,
$a^i_k\in \encode x$ and $f^i_\varsigma\in \encode x$;
moreover, as there is no action in $\Hits^{(1)}$ applicable in $\encode x$,
$h$ is applicable in $\encode x$:
$\encode x\PHPtrans \encode x\play h$.
In $\encode x\play h$, the only applicable actions of priority $1$ are those having
$a^i_{k'}$ as hitter and hitting cooperative sorts, giving a finite number of transitions towards
$\encode{x'}$.

(\ref{ph2adn}) $s\PHPtrans s'$ only if there exists an action $h$ applicable in $s$ such that
$s\play h=s'$.
If $\prio(h)=1$, then, by definition of $\Hits^{(1)}$, 
$\decode s=\decode {s'}$.
If $\prio(h)=2$, then $\forall i\in\segm{1}{n}$,
if $\get{s}{f^i} = f^i_\varsigma$, then, $\forall j\in\DNdep(f^i),
\get{\varsigma}{a^j}=\get{s}{a^j}$.
Let $i\in\segm{1}{n}$ such that $\get{s}{a^i}\neq\get{s'}{a^i}$ ($i$ is unique for this
transition).
By \pref{def:DN2PH}, if $\get{s'}{a^i}=a^i_{k'}$, necessarily $f^i(\decode s)=k'$, hence
$\decode s\DNtrans \decode{s'}$.
\end{proof}

>>>>

\section{Proof of the Weak Bisimulation of Asynchronous Discrete Networks (\pref{sec:dn})}
\label{suppl:proofbisimADN}

\newproof{proofbisimADN}{Proof of \pref{th:bisimDN}}
\begin{proofbisimADN}
(\ref{adn2ph}) From \pref{def:DN}, $x\DNtrans x'\Rightarrow \exists i\in\segm{1}{n},
f^i(x)=\get{x'}{i} \wedge \forall j\in\segm{1}{n},i\neq j, \get{x}{j}=\get{x'}{j}$.
Let us assume (without loss of generality) that $f^i(x)=k'$, $\get{x}{i}=k$ and
$\varsigma\in\underset{j\in\DNdep(f^i)}{\times} \PHl_{a^j}$ such that
$\forall j\in\DNdep(f^i), \get{\varsigma}{j}=a^j_{\get{x}{j}}$.
From \pref{def:DN2PH}, $h=\hit{f^i_\varsigma}{a^i_k}{a^i_{k'}}\in\Hits^{(2)}$.
From the definition of $\encode x$,
$a^i_k\in \encode x$ and $f^i_\varsigma\in \encode x$;
moreover, as there is no action in $\Hits^{(1)}$ applicable in $\encode x$,
$h$ is applicable in $\encode x$:
$\encode x\PHPtrans \encode x\play h$.
In $\encode x\play h$, the only applicable actions of priority $1$ are those having
$a^i_{k'}$ as hitter and hitting cooperative sorts, giving a finite number of transitions towards
$\encode{x'}$.

(\ref{ph2adn}) $s\PHPtrans s'$ only if there exists an action $h$ applicable in $s$ such that
$s\play h=s'$.
If $\prio(h)=1$, then, by definition of $\Hits^{(1)}$, 
$\decode s=\decode {s'}$.
If $\prio(h)=2$, then $\forall i\in\segm{1}{n}$,
if $\get{s}{f^i} = f^i_\varsigma$, then, $\forall j\in\DNdep(f^i),
\get{\varsigma}{a^j}=\get{s}{a^j}$.
Let $i\in\segm{1}{n}$ such that $\get{s}{a^i}\neq\get{s'}{a^i}$ ($i$ is unique for this
transition).
By \pref{def:DN2PH}, if $\get{s'}{a^i}=a^i_{k'}$, necessarily $f^i(\decode s)=k'$, hence
$\decode s\DNtrans \decode{s'}$.
\end{proofbisimADN}

### 2014-04-07 parts/suppl-ph-php.tex

\section{Proofs about flattening (\pref{ssec:flattening})}
\label{suppl:demoflattening}

>>>>

\section{Proof of the Weak Bisimulation of the Flattening (\pref{ssec:flattening})}
\label{suppl:demoflattening}

\newproof{proofbisimPHP}{Proof of \pref{th:bisimPHP}}
\begin{proofbisimPHP}
  (\ref{php2ph}) Let $\flats{\os} = s$.
    Given the dynamics of a PH (\pref{def:play}), if $\os \PHPtrans[\PH] \os'$,
    then there exists $h \in \ov{\PHh}$ playable in $\os$;
    therefore, $\os' = \os \PHplay h$.
    Thus, $\Fsem{\Fop{h}}{\os}$ and $\target{h} \in \os$,
    and given \pref{lem:ppplaysubset}, there exists $\mysigma \subseteq s$ so that $\Fsem{\Fop{h}}{\mysigma}$.
    Therefore, by construction of $\PH$ (\pref{def:flattening}) there exists
    $g = \PHhit{f^{h,i}_\mysigma}{\target{h}}{\bounce{h}} \in \PHh^{(2)}$.
    By construction of $s$ (\pref{def:flattening}), $\PHget{s}{f^{h,i}} = f^{h,i}_\mysigma$ and $\target{h} = \target{g} \in s$.
    Therefore, $g$ is playable in $s$.
    In $s \PHplay g$, the only playable actions are those in $\PHh^{(1)}$ having $\bounce{h} = \bounce{g}$ as hitter
    and updating cooperative sorts, allowing to reach $\flats{\os'}$ in a finite number of actions.
    Thus, $\flats{\os} \PHPtrans[\oPH]^* \flats{\os'}$.

  (\ref{ph2php}) Let $\os = \unflats{s}$.
    Given the dynamics of a PH (\pref{def:play}), if $s \PHPtrans s'$,
    then there exists $g \in \PHh$ playable in $s$; therefore, $s' = s \PHplay g$ and $\target{g} \in s$.
    If $\prio(g) = 1$ then only the active process of a cooperative sort has evolved, and $\unflats{s} = \unflats{s'}$.
    Otherwise, $\prio(g) = 2$; we note in this case: $g = \PHhit{f^{h,i}_\mysigma}{b_j}{b_k}$.
    By construction of the flattening (\pref{def:flattening}), there exists $h \in \ov{\PHh}$ so that
    $\target{h} = b_j$, $\bounce{h} = b_k$ and $\Fsem{\Fop{h}}{\mysigma}$.
    If $g$ is playable, this means that no other action in $\PHh^{(1)}$ is playable, and especially the cooperative sort
    $f^{h,i}$ is already updated; therefore, $\mysigma \subseteq \os$.
    Furthermore, $b_j \in \os$.
    Thus, by \pref{lem:ppplaysubset}, it comes that $h$ is playable in $\os$,
    and $\unflats{s} \PHPtrans \unflats{s'}$.
\end{proofbisimPHP}
